---
title: '"Pregnancy_Associated_Mortality_2020"'
author: "emmanuella_sainvilus"
date: "2024-02-20"
output: word_document
---

```{r setup, include=FALSE}
# Clear Memory
rm(list=ls())
gc()

# Set Working Directory
directory <- "C:/Users/emma1/OneDrive/NYC_Open_Project"
setwd(directory)


# Set Random Seed
set.seed(1432)

# Load Packages
library(readxl)
library(ggplot2)
```

```{r}
data <- read_excel("C:/Users/emma1/OneDrive/NYC_Open_Project/Pregnancy-Associated_Mortality_2020.xlsx")

data <- data.frame(data)
```



# Introduction

This analysis examines the relationship between the rate of maternal mortality in New York City and four different variables. These variables are borough, race/ethnicity, the number of these deaths determined to be related to the patient's pregnancy and finally the reported underlying causes of patient death

For this, we will use an NYC Open Data set looking at the maternal mortality rates of New York City reported for the year of 2020. 


## Analysis

## Bivariate Analysis

### Boroughs

This portion of the analysis observes the maternal mortality rates and compares them across different NYC boroughs. 

A separate dataset with the boroughs and their reported deaths is create so that R has less difficulty reading the numeric values of the variable. I do this for the other variables used in this analysis as well.  

For an unknown reason, each of the borough categories are duplicated in the data set. Despite this, each of the category's duplicates have a different number representing the number of deaths reported. For this reason, the number of deaths are added together only to their matching borough. This is done for ease of analysis. 

```{r}
Boroughs <- c("Bronx", "Brooklyn", "Queens", "Rest of State", "Manhattan", "Staten Island")
Deaths <- c(25, 22, 19, 8, 4, 2)



Boroughs_df <-  data.frame( Deaths = Deaths,
  Boroughs = factor(Boroughs, levels = Boroughs))

print(Boroughs_df)
```
```{r, include=FALSE}
summary(Boroughs_df)
```
To calculate the mean and standard deviation, a list is created only containing the numeric values of the "Boroughs" dataframe. This is done to avoid an R error, which the dataframe as a list of characters, and has difficulty pulling its numeric values. 

```{r,  include=FALSE}
Boroughs <- c(25, 22, 4, 19, 2, 8)
Boroughs_numeric <- as.numeric(Boroughs)
mean(Boroughs_numeric)
```
```{r,  include=FALSE}
sd(Boroughs_numeric)
```
The average number deaths calculated is about 13. Exactly half the boroughs reported having a death toll greater than the average. 

```{r}
Boroughs_df <- Boroughs_df[order(-Boroughs_df$Boroughs), ]

ggplot(Boroughs_df, aes(x = Boroughs, y = Deaths)) +
  geom_bar(stat = "identity", fill = "skyblue") + labs(title = "Maternal Deaths by Borough in 2020") + 
  ylim(0,30)
```

A bar graph comparing the number deaths for each category of the Boroughs varibable shows that the Bronx, Brooklyn and Queens have the highest maternal deaths, that are well above the average.


# Bivariate - Continous

# Related Variable

The Related variable refers to a "causal relationship between pregnancy and death categorized as pregnancy-related, pregnancy-associated but not rrelated or unable to determine preganancy-relatedness". In this portion of the analysis, we look to compare the number of deaths that are associated with, related to, and undetermined to be due to preganacy. 

```{r, include=FALSE}
Related <- c("Pregnancy-related", "Pregnancy-associated but not related", "Unable to Determine")

Deaths <- c( 87, 12, 10)
```

```{r, include=FALSE}
Related_df <-  data.frame(
 Deaths = Deaths,
  Related = factor(Related, levels = Related)
  )

print(Related_df)
```
```{r, include=FALSE}
summary(Related_df)
```
```{r, include=FALSE}
Related_numeric <- c(12, 87, 10)
sd(Related_numeric)

```
```{r, include=FALSE}
Related_numeric <- c(12, 87, 10)
mean(Related_numeric)
```
The average of maternal deaths calculated among the three categories of the Related varibable, is about 36. 
```{r}
ggplot(Related_df, aes(x = Related, y = Deaths)) +
  geom_bar(stat = "identity", fill = "skyblue") + labs(title = "Deaths Related to Pregnancy in 2020") + 
  ylim(0,110)
```


In a visual representation of the data, a bar graph demonstrates the pregnancy-related deaths were the highest reported of that year.




# Bivariate - Contiuous

## Race.ethnicity 

 This portion of the analysis observes the number of maternal deaths and compares them across different racial and ethnic groups. 

```{r, include=FALSE}
Race.ethnicity <- c("Black non-Latina", "Latina", "White non-Latina", "Asian/ Pacific Islander","Other")
numeric_values <- c(35, 22, 12, 10, 1)

Race.ethnicity_df <-  data.frame( Deaths = numeric_values,
  Race.ethnicity = factor(Race.ethnicity, levels = Race.ethnicity))

print(Race.ethnicity_df)
```
```{r, include=FALSE}
summary(Race.ethnicity_df)
```
```{r, include=FALSE}
Race.ethnicity_numeric <- c(12, 22, 35, 10, 1)
mean(Race.ethnicity_numeric)
```

The average amount of maternal deaths calculated for the Race/ethncity variable is 16. 

 Below is a bar graph, which provides a visual representation of the number of maternal deaths by race/                               
```{r}

ggplot(Race.ethnicity_df, aes(x = Race.ethnicity, y = Deaths)) +
  geom_bar(stat = "identity", fill = "skyblue") + labs(title = "Rates of Maternal Death by Race/Ethnicity in 2020") +
   ylim(0,40)
 
```

The graph output shows us that the racial/ethnic groups, "Black non-Latina" and "Latina" show death tolls nearly twice as high as the average for that year, while groups such as "White non-Latina" and "Asian/Pacific Islander" where below the average. 


## Bivariate - Continuous

### Underlying_Cause

This portion of the analysis observes the underlying_cause variable, which looks at the number of maternal deaths attributed to different underlying medical causes.

```{r, include=FALSE}
Underlying_cause <- c("Infection/Sepsis", "Mental Health Conditions (Overdose related to substance use disorder", "Other", "Hemorrhage", "Cardiovascular Conditions",  "Homicide", "Embolism", "Asthma/ pulmonary conditions", "Cancer", "Mental Health Conditions (Suicide)","Unknown COD", "Unintentional Injury")

numeric_values <- c(9, 9, 7, 6, 4, 4, 3, 2, 2, 2, 2, 1)

Underlying_df <-  data.frame( Deaths = numeric_values,
  Underlying_cause = factor(Underlying_cause, levels = Underlying_cause))

print(Underlying_df)
```
```{r, include=FALSE}
summary(Underlying_df)
```
```{r, include=FALSE}
Underlying_numeric <- c(5, 9, 2, 2, 6, 3, 2, 1, 7, 1, 1, 2)
mean(Underlying_numeric)
```

The average calculated from this dataset is about 3. The bar graph below shows us that the the four categories with attributed deaths higher than the calculated average are Mental Health Conditions (Suicide), Homicide, Mental Health Conditions (Overdose related to substance use disorder), and Infection/Sepsis with the highest reported death toll.


Below is a visual represenatation of the dataset looking at the relationship between number of deaths and types of underlying causes. 

```{r}
ggplot(Underlying_df, aes(x = Deaths, y = Underlying_cause)) +
  geom_bar(stat = "identity", fill = "skyblue") + labs(title = "Reported Causes of MMR In 2020") + xlim(0, 10)

```

